Cartogram Mapping and its Application to Cancer Data Visualisation

### 2019-09-24 review_article.Rmd
working on conclusion

---
title: "Cartogram Mapping and its Application to Cancer Data Visualization"
author: "Stephanie Kobakian, Jessie Roberts and Dianne Cook"
bibliography: reviewbib.bib
output:
  pdf_document:
    fig_caption: yes
    number_sections: yes
    toc: yes
    toc_depth: 3
  word_document:
    toc: yes
    toc_depth: '3'
link-citations: yes
fontsize: 11pt
biblio-style: vancouver
yolo: yes
---

@@

- Future steps should consider applications not currently utilized for cancer outcome displays, where their use may improve decision making
- needs to have a clear relevance to spatial patterns in cancer epidemiology
- presents alternative ways of visualizing those spatial patterns, or the uncertainty in those patterns
- a more applied-focused paper (e.g. lots of cool images), rather than theoretical (e.g. lots of formula).
-->

@@

# Introduction
<!-- What data are we working with, why, what are the purposes -->
Cancer statistics are usually delivered as an aggregated value for a geopolitical area. Presenting these statistics requires transforming individual observations into aggregations of communities as geographical units, in large part for privacy protection or political and policy purposes.  The information could be as simple as counts per area (e.g. state, province, local government area, post/zip code). Counts alone are not sufficient to compare areas, because the populations of areas are all different. In this case, the counts data needs to be merged with population data to appropriately calibrate it to incidence, for example, rate per 100000 people.  This type of data is collected on a routine basis for public health purposes, and may be made available to the general public as a service to the community. The task, then, is to examine what are the usual ways to communicate cancer statistics, to the public, are there alternative approaches, and what are the pros and cons of these choices.

<!-- Ways to visualize, whats our hypothesis -->
A common approach to communicate cancer statistics, is to display statistics on a map. Using a choropleth map: the statistic is mapped to color and the geographic region is filled with this color. The viewer would then be able to examine the spatial distribution of the disease incidence, where there is a trend in longitude or latitude, or rural vs urban, or coastal vs inland, or even specific hot spots of the disease. Visualizing diseases on maps is often the first step in exploratory spatial data analysis and effectively helps in the formulation of hypotheses [@MTMSIH]. Disease maps help to present geographic patterns that may be overlooked in a table, obscuring the geospatially related statistics [@SAMGIS].
By providing a visual representation of cancer outcomes, geographic patterns of disease may be identified and effectively addressed with public health policy and actions.
@SE recognizes one of the key challenges with mapping spatial patterns of disease is the design of visualizations.
This paper addresses the visualization techniques and their applications to cancer statistics. Highlighting the differences and historic use of these displays.

<!-- Structure of the paper -->
The paper is organized as follows. The next section describes the choropleth map which is the common approach to disease maps. Section \ref{ch:public} surveys atlases in use today. Section \ref{ch:alternatives} describes an alternative display, the cartogram what may be useful for countries that have heterogeneously sized geographic units. The pros and cons of these approaches is discussed in Section \ref{ch:compare}. Disease maps are more useful when made interactive, and common options are described in Section \ref{ch:interacting}, along with a discussion of benefits and disadvantages. The last section summarizes the paper and discusses future directions.

@@

<!-- what is a choropleth map -->
A choropleth map is used to display differences in the geographical distribution of data by spatial unit by shading areas of a map. The geography is faithfully rendered, and the color rendering is designed to reveal spatial patterns among data values. A choropleth is constructed by drawing the geographic or political boundaries, and filling the shapes with colors to represent values of a measured variable [@EI]. 
Figure \ref{fig:choroCRS} shows a choropleth of age-adjusted rate (per 100,000 people) of new cases of lung and bronchus in the USA, averaged over 2012 through 2016. The data was extracted from the official federal statistics by @usa_lung on cancer incidence and deaths, produced by the Centers for Disease Control and Prevention (CDC) and the National Cancer Institute (NCI).

<!-- history and types of choropleths, variations -->
Early versions of choropleth maps used symbols or patterns instead of color.
@CPISACA discuss the use of choropleths to visualize cancer data, and @DMAHP gives an overview of the development of these maps for displaying disease data.

<!-- psychology -->
Utilizing the state boundaries can make a map familiar to read [@CIBMUK], and allows viewers to visually infer the spatial relationships in the data. 
The familiarity of the geography is a worthy consideration when presenting results of spatial analysis. Just as geographers are no longer the only creators of maps, @CPISACA suggests the audiences of spatial health data analysis have extended beyond researchers to the public, policymakers and the media.

<!-- how do we use choropleth's -->
Identifying and explaining spatial structures, patterns, and processes involves considering the individuals in communities and organizing communities into representable units [@SAMGIS]. In Figure \ref{fig:choroCRS}, a west to east spatial trend of increasing rates, can be seen. There is also a spatial outlier -- Utah has a noticeably lower rate that its neighbors. Also Kentucky has a noticeably high rate, and Maine also has a higher rate than its neighbors. There is something of a cluster of higher rates around the tobacco states.

<!-- potential problems -->
While the areas are recognizable shapes, they are often politically driven boundaries with individual areas being of non-uniform size, containing different population densities and subject to change over time. 
The different population and geographical sizes of administrative areas can attract attention to the shades of the underpopulated but large areas [@EI], Skowronnek [-@BCM] calls this an area-size bias. 
Choropleths can inhibit visual inference when presenting human related statistics as the display may draw attention away from the 'potentially more important results in the more populous communities' that are geographically smaller [@SE].

```{r choropleth-grid-create, message=FALSE, warning = FALSE, fig.cap = "Publicly available choropleth cancer maps published between 2010 and 2015.", out.width='100%', fig.align='center', fig.width=12, fig.height=10}
library(cowplot)

@@

<!-- Displaying of cancer data on choropleths -->
Choropleth maps can be useful devices for communicating information to public on a familiar map base.
A cancer atlas is a choropleth map, or collection of maps, representing cancer incidence and mortality for a country, or group of countries. 
In epidemiology, choropleths are often used as a tool to study the spatial distribution of cancer incidence and mortality. 
The data collection methods of cancer mortality rates across regions, and the administrative control within regions lends itself to choropleth visualization. 
@MACM provides the definition of a cancer atlas, beginning with Haviland's maps in 1875, they attribute UK cancer atlases to Howe (1963), and early work in US cancer atlases can be attributed to Burbank (1971).
The increasing development and use of disease maps can be attributed to the availability of geographic information system software [@SE].
The choropleth maps presented levels via hatching or dots on a black and white scale.
These atlases were key to developing hypotheses regarding areas with unusually high rates, geographic correlations, work related exposures, and high risk diets [@MACM].

<!-- Measures reported in cancer atlases -->
The presentation of cancer statistics has changed over time with greater access to computational power.
Mortality rates are now often presented as relative rates of risk across the population, and age adjusted to correct for the higher prevalence of cancers in older populations.
Howe [-@HEDP] describes Stock's development of the standardized mortality ratios through the 1930s. Table \ref{tab:measures} summarizes the measures presented in published cancer atlases, and provides a definition of each measure.

@@

|Measure |Details| 
|:--------------------|:------------------------------------------------|  
|1. Count|Crude cancer counts|
|2. Rate per 100,000|Cancer incidence per 100,000 population|  
|3. New cancer cases per 100,000|Specific methods could not be found|   
|4. IR (Incidence Ratio)|$(IR)_i=\frac{(Incidence\ Rate)_i}{Average\ Incidence\ Rate}$, |
||Cancer incidence rate in region $i$ over the average cancer incidence rate for the total region|
|5. Age Adjusted Rate per 100,000| #5 standardized by age structure or region|   
|6. Age Adjusted Relative Risk| RR standardized by age structure in each region $i$|  
|7. SIR (Standardized Incidence Ratio)| IR standardized by age structure in each region $i$|
|8. Below or above Expected|Alternative expression of the SIR|
|9. RER |$RER = \frac{(Cancer\ related\ mortality)_i}{Average\ cancer\ related\ mortality}$|
|(Relative Excess Risk)|Represents the estimate of cancer related mortality within five years of diagnosis|
||Also referred to as 'excess hazard ratio'|

@@

Cancer maps are effective tools for communicating to wide range of audiences, including the general public and others not trained in statistical analyses.
These visualizations enable non-expert audiences to interpret the outputs of sophisticated statistical analyses. Cruickshank (1947) as cited by @DMAHP, discusses using visuals as a 'formal statistical assessment of the spatial pattern'. Overwhelmingly, cancer maps utilized to communicate to the public and other non-expert audiences are choropleths.  These atlases provided Incidence, Survival and Mortality rates to the public on areal or isopleth map bases.

The Atlas of Cancer in Queensland [@DACQ] focused on highlighting the difference in experience for those living in rural and disadvantaged areas, the Standardized Incidence Ratios (SIR) were modeled via Bayesian methods.
The presentation of these rates considers not only the shapes of the areas, but also appropriate choices of color blind friendly color schemes, and categories of the values to highlight significantly different areas.

@@

@roberts2019communication identified 33 publicly available cancer atlases, published between January 2010 and November 2015. All of these use choropleth maps. All but one of these were published by non-commercial organizations, including not-for-profits, government, research organizations, advocacy groups or government funded partnerships. The use of choropleths within the public domain mirrors the heavy use of choropleth maps within the research literature, discussed above. 
The cancer atlases identified by @roberts2019communication covered geographies from all around the world, most focused on single nations. Figure \ref{fig:choropleth-grid-create} displays a global (b), national (a, d, e, f) and state (c, g) choropleths. The sections below provide details on each of the maps within Figure \ref{fig:choropleth-grid-create}.

<!-- While there is significant variation across these maps in terms of the measure represented, the scale of the map, the color choices, the legend design and supplementary/supporting statistics, the mapping approach has not changes and almost exclusively only choropleth maps are used. Unlike innovation in spatial statistical methods, interactivity, data collection and record keeping little work has been done to implement different mapping approaches -->

@@

Figure 2.a contains an image from *[The Environmental and Health Atlas of England and Wales](http://www.envhealthatlas.co.uk/eha/Breast/)*. This map shows the relative risk for women developing lung cancer in England and Wales in 2010. The cancer data used to generate this map came from Office for National Statistics (ONS) (England) and from the Welsh Cancer Intelligence and Surveillance Unit (WCISU).

@@

The map seen in Figure 2.b is from the *[Globocan 2012: Estimated  Cancer Incidence, Mortality and Prevalence](http://globocan.iarc.fr/Pages/Map.aspx)*. This global map shows age standardized incidence rates (per 100,000) for all invasive cancers for both men and women, aggregated at a national level for 2018. This map is published by the [World Health organization's International Agency for Research on Cancer](https://www.iarc.fr/). Data was sourced from cancer registries in each country, contributing registries can be seen in the supplementary material on the cancer atlas website.

@@

Figure 2.c shows an extract from  *[The Atlas of Cancer in QLD](https://cancerqld.org.au/research/queensland-cancer-statistics/queensland-cancer-atlas/)*. This map was published by the Queensland Cancer Council and shows the relative incidence ratio of lung cancer in males in the state of QLD within Australia based on data from 1998 to 2007. Data to generate this map was sourced from the *[Queensland Cancer Registry](https://cancerqld.org.au/research/cancer-registries/queensland-cancer-registry/)*.

@@

One of the concerns of adding too much information to a map is the fear of cognitive overload [@mcgranaghan1993cartographic] in which the user reaches an information threshold, beyond which will not be able to make sense of the information. These concerns are not unfounded, complexity and density of representation methods appear to overwhelm novice decision makers, while experts are able to use the detail more readily when making-decisions [@cliburn2002design]. Interactivity is a design feature within modern mapping methods that can be used to incorporate additional information and complexity without overloading the user.

Interactivity enables supplementary information to be incorporated into online atlases without cluttering the display. Interactive design features found in online cancer maps include tool tip features, drop down menus, data selection, zooming and panning allow users to explore the map as they want more information and allow flexibility in the display [@HTLWM]. The use of these supports were found in various online cancer maps identified by @roberts2019communication. The controls for basic interactive features are often placed outside of the plot space [@TGA], thus the map image is updated/replaced as the user interacts with the controls. For example, changing the population age or other demographic variables. Some more advanced interactions include direct interactions with the plot via the use of overlaid tool tip features, very few cancer atlases involve these more complex selection interactive capabilities.

@@

The interactivity of modern mapping methods enables supplementary information to be incorporated without cluttering the screen.
Additionally, interactivity allows the user to toggle between different variables, map views or multiple realizations of possible future scenarios [@goodchild1994introduction]. Thus providing additional mechanisms for the users comprehension as well as the uncertainty of the available information (@maceachren1992visualizing; @van1994visualization).

These interactive features provide an opportunity for users to explore the additional information available.
Tool tip features and drop down menus allow for user exploration, small screens are not a problem with the flexibility provided by zooming and panning [@HTLWM]. This allows relationships between spatial areas and diseases to be explored with sophistication in nontraditional but still 'cognitively accessible' ways [@TNTEA].This helps users to understand and interpret the spatial distribution presented, as well as validate, explain or explore the presented statistics and their relationships to each other and/or their underlying spatial distribution. This allows relationships between spatial areas and diseases to be explored with sophistication in non-traditional but still 'cognitively accessible' ways [@TNTEA]. The interactive features of the publicly available maps identified by @roberts2019communication allow the exploration of geographic hierarchies, population distribution, statistical uncertainty, demographics and socio-economic indicators.
@TNTEA suggested LM plots as a solution to linking cartography and statistical graphics.

@@

Cancer maps are powerful visualizations that summarize complex statistical analyses, however the statistics represented in these maps cannot tell the entire story.
Supplementary graphs and plots are often included to add more depth and information to the map.
@CPISACA suggests additional materials such as tables, graphs, and text explanations support understanding and inference derived from maps, ensuring the message communicated will be consistent across a range of viewers.
There are many visualizations used for displays of statistical summaries, these may be dot plots, bar plots, box plots, times series plots, cumulative distribution plots, scatter plots, Q-Q plots. 
These additional displays of the cancer distribution can provide alternative views of the cancer statistics, as well as the supporting statistics including error, confidence intervals, distributions, sample or population sizes, standard deviation and other measures.
When presenting cancer maps, @MACM believes the intuition derived from maps must be 'validated by rigorous statistical analyses', the supplementary statistics help for this validation.

@@

While atlases are often used to describe differences between areas, statistics may be displayed at different levels of aggregation. Global health statistics can be aggregated to administrative and arbitrarily defined regions, such as those used by the World Health organization and the United Nations [@IARC_3].
World atlases can allow for displays of data aggregated into continents, countries, states, provinces and congressional districts [@USCS].

@@

It is extremely likely that each population area will have a different number of people. The distribution of the population residing in all areas may also be communicated in a table or histogram display [@NICR_1].
Atlases can connect the population to the land available to them by communicating population density.
Atlases can also connect the population to the land available to them by communicating population density.

### Statistical uncertainty
Additional statistics that accompany an atlas often include a measure of the statistical uncertainty surrounding the statistics presented in a choropleth.
In the review of atlases in the public domain, atlases were considered to report uncertainty to the non-expert user if they included a measure of statistical uncertainty either within or alongside the map [@roberts2019communication]. 
The maps considered used standard and well-known measures including credible intervals and standard deviation, statistical significance, box plots and distributions. Other methods involve providing adjacent maps or overlapping maps with symbols [@VSSDCUC].
The maps employing uncertainty ranged from static documents or infographics, to interactive online visualizations. Communicating the statistical uncertainty associated with the estimates occurs using confidence intervals (CI), credible intervals (CrI), statistical significance levels, box plots, distribution plots, and reporting sample size and standard deviations.
Close to half of the atlases identified (42%, n=14) included some measure of uncertainty. The most common measure used to represent uncertainty were credible or confidence intervals (CIs).

@@

### Socio-economic indicators
Socio-economic indicators can explain how the experience of cancer prevalence varies for various members of a society.
These indicators include unemployment rates, poverty rates, remoteness, and education levels achieved though, only a few atlases also explored the impact of rurality on cancer rates. 
These rates may also be explored as percentages above or below the mean or median value for the set of spatial areas.
The Human Development Index can be used to understand the socio-economic experience of a community, as well as Income levels which can be sourced from the World Bank list of economies [@IARC_3]. 
The areas are often ranked and allocated to quintiles; each quintile can be presented as categories describing the ranking.

One of the concerns of adding too much information to a map is the fear of cognitive overload [@mcgranaghan1993cartographic] in which the user reaches an information threshold, beyond which will not be able to make sense of the information. These concerns are not unfounded, complexity and density of representation methods appear to overwhelm novice decision makers, while experts are able to use the detail more readily when making-decisions [@cliburn2002design]. Interactivity is a design feature within modern mapping methods that can be used to incorporate additional information and complexity without overloading the user.

Interactivity enables supplementary information to be incorporated into online atlases without cluttering the display. Interactive design features found in online cancer maps include tool tip features, drop down menus, data selection, zooming and panning allow users to explore the map as they want more information and allow flexibility in the display [@HTLWM]. The use of these supports were found in various online cancer maps identified by @roberts2019communication. The controls for basic interactive features are often placed outside of the plot space [@TGA], thus the map image is updated/replaced as the user interacts with the controls. For example, changing the population age or other demographic variables. Some more advanced interactions include direct interactions with the plot via the use of overlaid tool tip features, very few cancer atlases involve these more complex selection tools.

>>>>

---
title: "Cartogram Mapping and its Application to Cancer Data Visualisation"
author: "Stephanie Kobakian, Jessie Roberts and Dianne Cook"
bibliography: reviewbib.bib
output:
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: 3
    fig_caption: true
fontsize: 11pt
link-citations: yes
biblio-style: vancouver
yolo: yes
---

@@

- Future steps should consider applications not currently utilised for cancer outcome displays, where their use may improve decision making
- needs to have a clear relevance to spatial patterns in cancer epidemiology
- presents alternative ways of visualising those spatial patterns, or the uncertainty in those patterns
- a more applied-focussed paper (e.g. lots of cool images), rather than theoretical (e.g. lots of formula).
-->

@@

# Introduction
<!-- What data are we working with, why, what are the purposes -->
Cancer statistics represent information on people, usually delivered as an aggregated value for a geopolitical area. Presenting statistics requires transforming individual observations into aggregations of communities as geographical units, in large part for privacy protection or political and policy purposes.  The information could be as simple as counts per area (e.g. state, province, local government area, post/zip code). Counts alone is not sufficient on which to compare different areas, because the populations of different areas are likely different. In this case, the counts data needs to be merged with population data to appropriately calibrate it to incidence, for example, rate per 100000 people.  This type of data is collected on a routine basis for public health purposes, and may be made available to the general public as a service to the community. The task, then, is to examine what are the usual ways to communicate cancer statistics, to the public, are there alternative approaches, and what are the pros and cons of these choices.

<!-- Geospatial statistics have been presented on maps to communicate information to the general public for centuries as they utilise familiar geographic shapes.
The mapping of diseases summarises spatial variations and can be considered a descriptive statistic of the relationship between a disease and the underlying geography [@SEMA].
The distribution of the statistic must be visible, it is not enough for areas on maps to be recognisable.
It is important to explore map views and designs that enhance the communication of the cancer statistics, especially as it is the populations that live on the land that are of interest, not the land itself. 
This has spurred innovations over the previous centuries to enhance maps to effectively communicate both cancer outcomes and health outcomes more broadly.-->

<!-- Ways to visualise, whats our hypothesis -->
A common approach to communicate cancer statistics, is to display statistics on a map. This is a choropleth map: the statistic is mapped to colour and the geographic region is filled with this colour. The viewer would then be able to examine the spatial distribution of the disease incidence, where there is a trend in longitude or latitude, or rural vs urban, or coastal vs inland, or even specific hotspots of the disease. Visualising diseases on maps is often the first step in exploratory spatial data analysis and effectively helps in the formulation of hypotheses [@MTMSIH]. Disease maps help to present geographic patterns that may be overlooked in a table, obscuring the geospatially related statistics [@SAMGIS].
By providing a visual representation of cancer outcomes, geographic patterns of disease are able to be identified and effectively addressed with public healh policy and actions.
@SE recognises one of the key challenges with mapping spatial patterns of disease is the design of visualisations.
This paper addresses the visualisation techniques and their applications to cancer statistics. Highlighting the differences and historic use of these displays.

<!-- Structure of the paper -->
The paper is organised as follows. The next section describes the choropleth map which is the common approach to disease maps. Section \ref{ch:public} surveys atlases in use today. Section \ref{ch:alternatives} describes an alternative display, the cartogram what may be useful for countries that have heterogeneously sized geographic units. The pros and cons of these approaches is discussed in Section \ref{ch:compare}. Disease maps are more useful when made interactive, and common options are described in Section \ref{ch:interacting}, along with a discussion of benefits and disadvantages. The last section summarises the paper and discusses future directions.

@@

<!-- what is a choropleth map -->
A choropleth map is used to display differences in the geographical distribution of data by spatial unit by shading areas of a map. The geography is faithfully rendered, and the colour rendering is designed to reveal spatial patterns among data values. A choropleth is constructed by drawing the geographic or political boundaries, and filling the shapes with colours to represent values of a measured variable [@EI]. 
Figure \ref{fig:choroCRS} shows a choropleth of age-adjusted rate (per 100,000 people) of new cases of lung and bronchus in the USA, averaged over 2012 through 2016. The data was extracted from the official federal statistics by @usa_lung on cancer incidence and deaths, produced by the Centers for Disease Control and Prevention (CDC) and the National Cancer Institute (NCI).

<!-- history and types of choropleths, variations -->
Early versions of choropleth maps used symbols or patterns instead of colour.
@CPISACA discuss the use of choropleths to visualise cancer data, and @DMAHP gives an overview of the development of these maps for displaying disease data.

<!-- psychology -->
Utilising the state boundaries can make a map familiar to read [@CIBMUK], and allows viewers to visually infer the spatial relationships in the data. 
The familiarity of the geography is a worthy consideration when presenting results of spatial analysis. Just as geographers are no longer the only creators of maps, @CPISACA suggests the audiences of spatial health data analysis have extended beyond researchers to the public, policymakers and the media.

<!-- how do we use choropleth's -->
Identifying and explaining spatial structures, patterns, and processes involves considering the individuals in communities and organising communities into representable units [@SAMGIS]. In Figure \ref{fig:choroCRS}, a west to east spatial trend of increasing rates, can be seen. There is also a spatial outlier -- Utah has a noticeably lower rate that its neighbours. Also Kentucky has a noticeably high rate, and Maine also has a higher rate than its neighbours. There is something of a cluster of higher rates around the tobacco states.

<!-- potential problems -->
While the areas are recognisable shapes, they are often politically driven boundaries with individual areas being of non-uniform size, containing different population densities and subject to change over time. 
The different population and geographical sizes of administrative areas can attract attention to the shades of the unpopulated but large areas [@EI], Skowronnek [-@BCM] calls this an area-size bias. 
Choropleths can inhibit visual inference when presenting human related statistics as the display may draw attention away from the 'potentially more important results in the more populous communities' that are geographically smaller [@SE].

```{r choropleth-grid-create, message=FALSE, warning = FALSE, fig.cap = "Publicly available choropleth cancer maps, published between 2010 and 2015.", out.width='100%', fig.align='center', fig.width=12, fig.height=10}
library(cowplot)

@@

<!-- Displaying of cancer data on choropleths -->
Choropleth maps can be useful devices for communicating information to public on a familiar map base.
A cancer atlas is a choropleth map, or collection of maps, representing cancer incidence and mortality for a country, or group of countries. 
In epidemiology, choropleths are often used as a tool to study the spatial distribution of cancer incidence and mortality. 
The data collection methods of cancer mortality rates across regions, and the administrative control within regions lends itself to choropleth visualisation. 
@MACM provides the definition of a cancer atlas, beginning with Haviland's maps in 1875, they attribute UK cancer atlases to Howe (1963), and early work in US cancer atlases can be attributed to Burbank (1971).
The increasing development and use of disease maps can be attributed to the availability of geographic information system software [@SE].
The choropleth maps presented levels via hatchings or dots on a black and white scale.
These atlases were key to developing hypotheses regarding areas with unusually high rates, geographic correlations, work related exposures, and high risk diets [@MACM].

<!-- Measures reported in cancer atlases -->
The presentation of cancer statistics has changed over time with greater access to computational power.
Mortality rates are now often presented as relative rates of risk across the population, and age adjusted to correct for the higher prevalence of cancers in older populations.
Howe [-@HEDP] describes Stock's development of the standardised mortality ratios through the 1930s. Table \ref{tab:measures} summarises the measures presented in published cancer atlases, and provides a definition of each measure.

@@

|Measure |Details| 
|:--------------------|:------------------------------------------------|  
|1. Count|Crude cancer counts|
|2. Rate per 100,000|Cancer incidence per 100,000 population|  
|3. New cancer cases per 100,000|Specific methods could not be found|   
|4. IR (Incidence Ratio)|$(IR)_i=\frac{(Incidence\ Rate)_i}{Average\ Incidence\ Rate}$, |
||Cancer incidence rate in region $i$ over the average cancer incidence rate for the total region|
|5. Age Adjusted Rate per 100,000| #5 standardised by age structure or region|   
|6. Age Adjusted Relative Risk| RR standardised by age structure in each region $i$|  
|7. SIR (Standardised Incidence Ratio)| IR standardised by age structure in each region $i$|
|8. Below or above Expected|Alternative expression of the SIR|
|9. RER |$RER = \frac{(Cancer\ related\ mortality)_i}{Average\ cancer\ related\ mortality}$|
|(Relative Excess Risk)|Represents the estimate of cancer related mortality within five years of diagnosis|
||Also referred to as 'excess hazard ratio'|

@@

The atlas of Cancer in Queensland [@DACQ] focussed on highlighting the difference in experience for those living in rural and disadvantaged areas. They used Standardised Incidence Ratios modelled via Bayesian methods, and explored the information presented in a range of selected atlases. These atlases provided Incidence, Survival and Mortality rates to the public on Areal or Isopleth map bases.

The presentation of these rates considers not only the shapes of the areas, but also appropriate choices of colour blind friendly colour schemes, and categories of the values to highlight significantly different areas.
<!--

Examples of atlases:
Kraak 1998, Kraak and Ormeling 2011
Bertin 1967
Coleman MP, Gatta G, Verdecchia A, Esteve J, Sant M, Storm H, Allemani C,Ciccolallo L, Santaquilani M, Berrino F:EUROCARE-3 summary: cancersurvival in Europe at the end of the 20th century.Ann Oncol2003,14(Suppl 5):v128-149

National Cancer Institute:Cancer Trends Progress Report - 2009/2010 UpdateBethesda: NCI, NIH, DHHS; 2010

Pickle LW:A history and critique of U.S. mortality atlases.Spatial andSpatio-temporal Epidemiology2009,1:3-17.

Mason TJ, McKay FW, Hoover R, Blot WJ, Fraumeni JF:Atlas of cancermortality for U.S. counties: 1950-1969Washington: U.S. Govt. Printing Office;1975.
howmick T, Robinson AC, Gruver A, MacEachren AM, Lengerich EJ:Distributed usability evaluation of the Pennsylvania Cancer Atlas.Int JHealth Geogr2008,7:36.

Queensland Cancer Registry:Cancer in Queensland: Incidence, Mortality,Survival and Prevalence,1982 to 2007Brisbane: QCR, Cancer CouncilQueensland and Queensland Health; 2010.

New York State Department of Health Maps of Cancer Incidence byCounty.[http://www.health.state.ny.us/statistics/cancer/registry/cntymaps/index.htm].
Quinn M, Wood H, Cooper N, Rowan S, Eds:Cancer Atlas of the UnitedKingdom and Ireland 1991-2000.London: Office for National Statistics;2005.
-->

Cancer maps are effective tools for communicating to wide range of audiences, including the general public and others not trained in statistical analyses.
These visualisations enable non-expert audiences to interpret the outputs of sophisticated statistical analyses. Cruickshank (1947) as cited by @DMAHP, discusses using visuals as a 'formal statistical assessment of the spatial pattern'. Overwhelmingly, cancer maps utilised to communicate to the public and other non-expert audiences are choropleths.

@@

@roberts2019communication identified 33 publicly available cancer atlases, published between January 2010 and November 2015. All of these use choropleth maps. All but one of these were published by non-commercial organisations, including not-for-profits, government, research organisations, advocacy groups or government funded partnerships. The use of choropleths within the public domain mirrors the heavy use of choropleth maps within the research literature, discussed above. 
The cancer atlases identified by @roberts2019communication covered geographies from all around the world, most focused on single nations. Figure \ref{fig:choropleth-grid-create} displays a global (b), national (a,d,e,f) and state (c,g) choropleths. The sections below provide details on each of the maps within Figure \ref{fig:choropleth-grid-create}.

<!-- While there is significant variation across these maps in terms of the measure represented, the scale of the map, the colour choices, the legend design and supplementary/supporting statistics, the mapping approach has not changes and almost exclusively only choropleth maps are used. Unlike innovation in spatial statistical methods, interactity, data collection and record keeping little work has been done to implement different mapping approaches -->

@@

Figure 2.a contains an image from the *[The Environmental and Health Atlas of England and Wales](http://www.envhealthatlas.co.uk/eha/Breast/)*. This map shows the relative risk for women developing lung cancer in England and Wales in 2010. The cancer data used to generate this map came from Office for National Statistics (ONS) (England) and from the Welsh Cancer Intelligence and Surveillance Unit (WCISU).

@@

The map seen in Figure 2.b is from the *[Globocan 2012: Estimated  Cancer Incidence, Mortality and Prevalence](http://globocan.iarc.fr/Pages/Map.aspx)*. This global map shows age standardized incidence rates (per 100,000) for all invasive cancers for both men and women, aggregated at a national level for 2018. This map is published by the [World Health Organisation's International Agency for Research on Cancer](https://www.iarc.fr/). Data was sourced from cancer registries in each country, contributing registries can be seen in the supplementary material on the cancer atlas website.

@@

Figure 2.c shows an extract from the *[The Atlas of Cancer in QLD](https://cancerqld.org.au/research/queensland-cancer-statistics/queensland-cancer-atlas/)*. This map was published by the Queensland Cancer Council and shows the relative incidence ratio of lung cancer in males in the state of QLD within Australia based on data from 1998 to 2007. Data to generate this map was sourced from the *[Queensland Cancer Registry](https://cancerqld.org.au/research/cancer-registries/queensland-cancer-registry/)*.

@@

One of the concerns of adding too much information to a map is the fear of cognitive overload [@mcgranaghan1993cartographic] in which the the user reaches an information threshold, beyond which will not be able to make sense of the information. These concerns are not unfounded, complexity and density of representation methods appear to overwhelm novice decision makers, while experts are able to use the detail more readily when making-decisions [@cliburn2002design]. Interactivity is a design feature within modern mapping methods that can be used to incorporate additional information and complexity without overloading the user.

Interactivity enables supplementary information to be incorporated into online atlases without cluttering the display. Interactive design features found in online cancer maps include tool tip features, drop down menus, data selection, zooming and panning allow users to explore the map as they want more information and allow flexibility in the display [@HTLWM]. The use of these supports were found in various online cancer maps identified by @roberts2019communication. The controls for basic interactive features are often placed outside of the plot space [@TGA], thus the map image is updated/replaced as the user interacts with the controls. For example, changing the population age or other demographic variables. Some more advanced interactions include direct interactions with the plot via the use of overlaid tool tip features, very few cancer atlases involve these more complex selection interactivities.

@@

The interactivity of modern mapping methods enable supplementary information to be incorporated without cluttering the screen.
Additionally, interactivity allows the user to toggle between different variables, map views or multiple realizations of possible future scenarios [@goodchild1994introduction]. Thus providing additional mechanisms for the users comprehension as well as the uncertainty of the available information (@maceachren1992visualizing; @van1994visualization).

These interactive features provide an opportunity for users to explore the additional information available.
Tool tip features and drop down menus allow for user exploration, small screens are not a problem with the flexibility provided by zooming and panning [@HTLWM]. This allows relationships between spatial areas and diseases to be explored with sophistication in nontraditional but still 'cognitively accesible' ways [@TNTEA].This helps users to understand and interpret the spatial distribution presented, as well as validate, explain or explore the presented statistics and their relationships to each other and/or their underlying spatial distribution. This allows relationships between spatial areas and diseases to be explored with sophistication in non-traditional but still 'cognitively accesible' ways [@TNTEA]. The interactive features of the publically available maps identified by @roberts2019communication allow the exploration of geographic hierarchies, population distribution, statistical uncertainty, demographics and socio-economic indicators.
@TNTEA suggested LM plots as a solution to linking cartography and statistical graphics.

@@

Cancer maps are powerful visualisations that summarise complex statistical analyses, however the statistics represented in these maps cannot tell the entire story.
Supplementary graphs and plots are often included to add more depth and information to the map.
@CPISACA suggests additional materials such as tables, graphs, and text explanations support understanding and inference derived from maps, ensuring the message communicated will be consistent across a range of viewers.
There are many visualisations used for displays of statistical summaries, these may be dot plots, bar plots, box plots, timesseries plots, cumulative distribution plots, scatter plots, Q-Q plots. 
These additional displays of the cancer distribution can provide alternative views of the cancer statistics, as well as the supporting statistics including error, confidence intervals, distributions, sample or population sizes, standard deviation and other measures.
When presenting cancer maps, @MACM believes the intuition derived from maps must be 'validated by rigorous statistical analyses', the supplementary statistics help for this validation.

@@

While atlases are often used to describe differences between areas, statistics may be displayed at different levels of aggregation. Global health statistics can be aggregated to administrative and arbitrarily defined regions, such as those used by the World Health Organisation and the United Nations [@IARC_3].
World atlases can allow for displays of data aggregated into continents, countries, states, provinces and congressional districts [@USCS].

@@

It is extremely likely that each population area will have a different number of people. The distribution of the population residing in all areas may also be communicated in a table or histogram display [@NICR_1].
Atlases have the opportunity to connect the population to the land available to them by communicating population density.
Atlases can also connect the population to the land available to them by communicating population density.

### Statistical uncertainty
Additional statistics that accompany an atlas often include a measure of the statistical uncertainty surrounding the statistics presented in a choropleth.
In the review of atlases in the public domain, atlases were considered to report uncertainty to the non-expert user if they included a measure of statistical uncertainty either within or alongside the map [@roberts2019communication]. 
The maps considered used standard and well known measures including credible intervals and standard deviation, statistical significance, box plots and distributions. Other methods involve providing adjacent maps or overlapping maps with symbols [@VSSDCUC].
The maps employing uncertainty ranged from static documents or infographics, to interactive online visualisations. Communicating the statistical uncertainty associated with the estimates occurs using confidence intervals (CI), credible intervals (CrI), statistical significance levels, boxplots, distribution plots, and reporting sample size and standard deviations.
Close to half of the atlases identified (42%, n=14) included some measure of uncertainty. The most common measure used to represent uncertainty were credible or confidence intervals (CIs).

@@

### Socio-economic indicators
Socio-economic indicators can explain how the experience of cancer prevalence varies for various members of a society.
These indicators include unemployment rates, poverty rates, remoteness, and education levels achieved though, only a few atlases also explored the impact of rurality on cancer rates. 
These rates may also be explored as percentages above or below the mean or median value for the set of spatial areas.
The Human Development Index can be used to understand the socio-economic experience of a community, as well as Income levels which can be sourced from the World Bank list of economies [@IARC_3]. 
The areas are often ranked and allocated to quintiles, each quintile can be presented as categories describing the ranking.

One of the concerns of adding too much information to a map is the fear of cognitive overload [@mcgranaghan1993cartographic] in which the the user reaches an information threshold, beyond which will not be able to make sense of the information. These concerns are not unfounded, complexity and density of representation methods appear to overwhelm novice decision makers, while experts are able to use the detail more readily when making-decisions [@cliburn2002design]. Interactivity is a design feature within modern mapping methods that can be used to incorporate additional information and complexity without overloading the user.

Interactivity enables supplementary information to be incorporated into online atlases without cluttering the display. Interactive design features found in online cancer maps include tool tip features, drop down menus, data selection, zooming and panning allow users to explore the map as they want more information and allow flexibility in the display [@HTLWM]. The use of these supports were found in various online cancer maps identified by @roberts2019communication. The controls for basic interactive features are often placed outside of the plot space [@TGA], thus the map image is updated/replaced as the user interacts with the controls. For example, changing the population age or other demographic variables. Some more advanced interactions include direct interactions with the plot via the use of overlaid tool tip features, very few cancer atlases involve these more complex selection interactivities.
